Concurrent Programming with CXL

### 2020-06-21 book/paper.tex
Added another exercise
Changed definition of blocking and added some exercises

  description={is a process cannot make progress without the help of another process.

@@

      application-defined condition while changing the state, for example by
      continually acquiring and releasing a lock}

@@

While the CVM does not have any specific built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
atomic_inc(&count);
\end{Verbatim}
\end{code}

@@

We call a process \emph{blocked}
\index{blocked process}
\glsadd{blocked process}
in a state if the process is in
an \emph{infinite loop} in which the process does not update the
shared memory.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.
The only way to break out of the infinite loop is if another process
changes the shared memory.
The \texttt{while} loop in Peterson's algorithm, in case its condition
does not hold, is another example of a process being blocked.

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
\glsadd{busy waiting}
(aka spin-waiting):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

However, there is a big difference.  As we pointed out, the processes that
are waiting for a spinlock are \emph{blocked}: they are in an infinite loop
in which they do not change the state.
The processes waiting for a reader/writer lock do change the state while
in the infinite loop: they acquire the mutex and release it.

In most operating systems and programming language runtimes,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
CXL supports this with the \texttt{synchS} module.
Busy waiting disables all this: even when using the \texttt{synchS}
module, readers and writers only get suspended temporarily in case
there is contention for the mutex.
A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
until the read or write lock is available.

@@

\section*{Exercises}
\begin{enumerate}
\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
per process rather than a simple counter?
Can you write a version of this test that uses a counter instead
of a flag per process?
% \item Write a ``library implementation'' of reader/writer locks so multiple
% reader/writer locks can be instantiated.
\item Write a test program for ``bound locks'' (as implemented in the
exercises of the last chapter) that checks that up to \texttt{M} processes
can acquire the bound lock at the same time.
\item Suppose a sports club has only one dressing room.  The policy is that
either only women or only men may be in the dressing room at any time.
Implement a ``dressing room lock'' that allows multiple people to enter the
dressing room while observing the policy.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Does it matter in what order the philosophers lay down their forks?
\end{enumerate}

@@

@@

\section*{Exercises}
\begin{enumerate}
\item Assume that you have only binary semaphores.  Implement counting
semaphores using binary semaphores using split binary semaphores.
\item Implement a solution to the producer/consumer problem
(Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
\end{enumerate}

>>>>

  description={is a process that is waiting on a low-level synchronization primitive
    and cannot make progress without the help of another process.

@@

      application-defined condition}

@@

While the CVM does not have any specific built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} statement blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
atomic_inc(&count);
\end{Verbatim}
\end{code}

@@

\glsadd{blocked process}
We call a process \emph{blocked}
\index{blocked process}
in a state if the process is waiting on a low-level synchronization
variable and cannot make progress without the help of another process.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked.
The only way forward is if the other process releases the lock.

@@

\glsadd{busy waiting}
While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
(aka spin-waiting):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
Figure~\ref{fig:inc} also has an example of busy-waiting: the \texttt{main}
process waits for the other two processes to finish by checking their \texttt{done}
flags.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

However, there is a big difference.  As we pointed out, the processes that
are waiting for a spinlock are \emph{blocked}.
In most operating systems and programming language runtimes,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
CXL supports this with the \texttt{synchS} module.
Busy waiting disables all this: even when using the \texttt{synchS}
module, readers and writers only get suspended temporarily in case
there is contention for the mutex.
A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
until the read or write lock is available.

@@

\section*{Exercises}
\begin{enumerate}
\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
per process rather than a simple counter?
Can you write a version of this test that uses a counter instead
of a flag per process?
% \item Write a ``library implementation'' of reader/writer locks so multiple
% reader/writer locks can be instantiated.
\item Write a test program for ``bound locks'' (as implemented in the
exercises of the last chapter) that checks that up to \texttt{M} processes
can acquire the bound lock at the same time.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Does it matter in what order the philosophers lay down their forks?
\item Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\end{enumerate}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/bank.cxl}
\end{code}
\caption{[bank.cxl] Bank accounts.}
\label{fig:bank}
\end{figure}

@@

\section*{Exercises}
\begin{enumerate}
\item Change the code in Figure~\ref{fig:boundedbuffer} so that
\texttt{produce()} fails if the buffer is full rather than block.
\texttt{produce()} should return \texttt{False} if it fails
and return \texttt{True} if it succeeds.
\item Implement a \emph{bounded stack}.  Method \texttt{push(item)}
should block if the stack is full.  Method \texttt{pop()} should block
until the stack is non-empty and then return an item.
\item Implement a GPU allocator.  There are \texttt{N} GPUs identified
by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease(gpu)} releases the given GPU.  It never needs
to block.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Assume that you have only binary semaphores.  Implement counting
semaphores using binary semaphores using split binary semaphores.
\item Implement a solution to the producer/consumer problem
(Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
\item Suppose a sports club has only one dressing room.  The policy is that
either only women or only men may be in the dressing room together.
Implement a ``dressing room lock'' that allows multiple people to enter the
dressing room while observing the policy.
\end{enumerate}
